\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-06 paper.tex

>>>>

\subsection{False positive recombinants}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to large numbers of samples, and are almost certainly
artefactual. Table~\ref{tab:false_positive} shows details for four recombinants
in the Long ARG that are likely false positives, and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869),
labelled A--D. These false positives are associated with the origins
of large outbreaks. For each row, we show the sample's Pango lineage
designation and the rank of that sample in the overall list of
samples with that designation. Thus, rows C and D are the first
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are the third and eighth samples from the B.1.617.2
lineage (Delta VoC). Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of 3; Section~\ref{sec:ls}).
The column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

\begin{table} \centering
\begin{tabular}{llllll}
\toprule
& Strain & Descendants & Lineage & Rank in lineage&HMM Cost \\
\midrule
A&Germany/HH-RKI-I-061284/2021 & 178,405    & B.1.617.2   & 3& 17  \\
B&India/ILSGS00961/2021        & 177,649    & B.1.617.2   & 8& 25   \\
C&Denmark/DCGC-281594/2021     & 127,227    & BA.2.9      & 1& 32 \\
D&USA/NJ-GBW-EWR000001/2021    & 127,230    & BA.3        & 1& 25 \\
\bottomrule
\end{tabular}
\caption{\label{tab:false_positive}
False positive recombination events. Details show are for the top
four recombination nodes in the Long ARG ordered by number of
descending samples. See the text for details on the remaining columns.
Rows are labelled A, B, C and D for ease of reference.}
\end{table}

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps
is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterisation of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will match in to
the existing ARG with very low likelihood, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
in segments. If this is combined with data quality issues and
``time travellers'' (Section~\ref{sec:filtering_time_travellers}),
the early stages of the VoC emergence can be quite muddled.
This appears to be the case for the Delta VoC,
which corresponded to the sudden emergence of
[HOW MANY, citation?] mutations,
as show in Table~\ref{tab:false_positive}A and B.
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

Given these considerations, it  is important to note that the
number of ultimate descendants is not an entirely reliable
indicator of reliability for recombinants.
Further study is required to systematically identify such false
positives recombinants, and to update the topology with a
more parsimonious explanation of the data.
